---
title: "New R Package deltatest: Statistical Hypothesis Testing Using the Delta Method for Online A/B Testing"
author: "Koji Makiyama"
date: "2025-03-15"
categories: [R]
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

In online A/B testing, we often face a significant practical challenge: the randomization unit differs from the analysis unit. Typically, control and treatment groups are randomly assigned at the user level, while metrics—such as click-through rate—are measured at a more granular level (e.g., per page-view). In this case, the randomization unit is user, but the analysis unit is page-view.

This discrepancy raises concerns for statistical hypothesis testing, which assumes that data points are independent and identically distributed (i.i.d.). Specifically, a single user can generate multiple page-views, and each user may have a different probability of clicking. Consequently, the data may exhibit within-user correlation, thereby violating the i.i.d. assumption.

When the standard Z-test is applied to such correlated data, the resulting p-values do not follow the expected uniform distribution under the null hypothesis. As a result, smaller p-values tend to occur more frequently even when there is no true difference, increasing the risk of falsely detecting a significant difference.

![](20250305012135)

To address this problem, Deng et al. (2018) proposed a modified statistical hypothesis testing method. Their approach replaces the standard variance estimation formula in the Z-test with an approximate formula derived via the Delta method, which accounts for within-user correlation. To simplify the application of this method, the **deltatest** package has been developed.

To illustrate how to use this package, we prepare a data frame that includes columns for the number of clicks and page-views aggregated for each user. This data frame also contains a column indicating whether each user was assigned to the control or treatment group.

```{r}
library(dplyr)

n_user <- 2000

set.seed(314)
data <- deltatest::generate_dummy_data(n_user) |> 
  mutate(group = if_else(group == 0, "control", "treatment")) |>
  group_by(user_id, group) |> 
  summarise(clicks = sum(metric), pageviews = n(), .groups = "drop")

data
```

The statistical hypothesis test using the Delta method can then be performed on this data as follows:

```{r}
library(deltatest)

deltatest(data, clicks / pageviews, by = group)
```

This version of the Z-test yields p-values that follow the expected uniform distribution under the null hypothesis, even when within-user correlation is present.

![](20250305015654)

For more details, refer to [https://github.com/hoxo-m/deltatest](https://github.com/hoxo-m/deltatest).

## Installation

You can install the **deltatest** package from CRAN.

```r
install.packages("deltatest")
```

## References

- [CRAN: Package deltatest](https://cran.r-project.org/package=deltatest)
- [hoxo-m/deltatest: Statistical Hypothesis Testing Using the Delta Method for Online A/B Testing](https://github.com/hoxo-m/deltatest)
- Deng, A., Knoblich, U., & Lu, J. (2018). Applying the Delta Method in Metric Analytics: A Practical Guide with Novel Ideas. *Proceedings of the 24th ACM SIGKDD International Conference on Knowledge Discovery & Data Mining.* [doi:10.1145/3219819.3219919](https://doi.org/10.1145/3219819.3219919)
