---
title: "deltatest 0.2.0: Statistical Hypothesis Testing Using the Delta Method for Online A/B Testing"
author: "Koji Makiyama"
date: 2026-3-12
categories: [R]
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 4,
  fig.height = 3
)
set.seed(314)
```

I’m happy to share a new release of **deltatest**.

This release includes two main changes:

- a new `tidy()` method for `deltatest` objects
- a fix for p-value calculation in one-sided tests

Before looking at what changed in this release, let’s briefly revisit the purpose of **deltatest**.

## What deltatest is for

The **deltatest** package provides `deltatest()`, a function for performing two-sample Z-tests using the delta method.

It is designed for common settings in online A/B testing where:

- randomization is done at the user level, but
- the metric is measured at a finer unit such as page views or sessions.

In such settings, naive tests can underestimate uncertainty—for example, standard Z-tests, chi-squared tests, or tests for differences in proportions—because observations within a user are not independent. `deltatest()` addresses this issue by using a delta-method-based variance estimator.

```r
# Install the released version from CRAN
install.packages("deltatest")

# Load packages
library(dplyr)
library(deltatest)

# Generate dummy data
data <- deltatest::generate_dummy_data(2000) |>
  mutate(group = if_else(group == 0, "control", "treatment")) |>
  group_by(user_id, group) |>
  summarise(clicks = sum(metric), pageviews = n(), .groups = "drop")

# Run a test
deltatest(data, clicks / pageviews, by = group)
```

Typical output:

```r
#> Two Sample Z-test Using the Delta Method
#> 
#> data:  clicks/pageviews by group
#> Z = 0.31437, p-value = 0.7532
#> alternative hypothesis: true difference in means between control and treatment is not equal to 0
#> 95 percent confidence interval:
#>  -0.01410593  0.01949536
#> sample estimates:
#>   mean in control mean in treatment        difference
#>       0.245959325       0.248654038       0.002694713
```

## What’s new in 0.2.0

### `tidy()` support for `deltatest` objects

With this release, `deltatest()` results can now be converted directly into a tidy tibble with `broom::tidy()`.

`deltatest()` returns an `htest`-class object, which is convenient for printing and interactive use. But in a tidyverse workflow, it is often much easier to work with results in a tidy tibble format. This makes it easier to combine results across many experiments or metrics, and to visualize patterns in estimates, confidence intervals, or p-values with tools like **ggplot2**.

First, here is a simple example of converting the result to a tidy format:

```{r tidy-example1}
library(dplyr)
library(deltatest)
library(broom)

data <- deltatest::generate_dummy_data(2000) |>
  mutate(group = if_else(group == 0, "control", "treatment")) |>
  group_by(user_id, group) |>
  summarise(clicks = sum(metric), pageviews = n(), .groups = "drop")

result <- deltatest(data, clicks / pageviews, by = group)

tidy(result)
```

Next, here is an example of using the tidy results to compare multiple experiments in a plot:

```{r tidy-example2}
library(ggplot2)

data2 <- deltatest::generate_dummy_data(2000, xi = 0.05) |>
  mutate(group = if_else(group == 0, "control", "treatment")) |>
  group_by(user_id, group) |>
  summarise(clicks = sum(metric), pageviews = n(), .groups = "drop")

result2 <- deltatest(data2, clicks / pageviews, by = group)

result_tidy1 <- tidy(result)  |> mutate(experiment_id = "test01")
result_tidy2 <- tidy(result2) |> mutate(experiment_id = "test02")

result_tidy <- bind_rows(result_tidy1, result_tidy2)

ggplot(result_tidy, aes(experiment_id, estimate)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, color = "red") +
  xlab(NULL) + ylab("Estimated CTR difference") +
  ggtitle("Treatment effects by experiment")
```

### Fix for one-sided p-value calculation

This release also fixes a bug in the p-value calculation for one-sided tests. In the previous version, p-values for one-sided tests could be incorrectly calculated using the two-sided formula. That behavior has now been fixed.

I would like to thank Kazuyuki Sano for reporting this issue and contributing to the fix.

## Final thoughts

I’m glad to keep improving **deltatest** little by little. If you use R for online A/B experiments, I hope it is useful to you.

For more details, see:

- Package website: [https://hoxo-m.github.io/deltatest/](https://hoxo-m.github.io/deltatest/)
- GitHub repository: [https://github.com/hoxo-m/deltatest](https://github.com/hoxo-m/deltatest)
